---
title: "Lab 1"
author: |
  | Bahaa Khaled Mohamed 21010383
  | Omar Aldawy Ibrahim Aldawy 21010864
date: "`r Sys.Date()`"
output:
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
    highlight: tango
    fig_caption: true
  html_document:
    df_print: paged
geometry: margin=1in
linestretch: 1.5
header-includes:
  - \usepackage{fancyhdr}
  - \usepackage{framed}
  - \usepackage{tcolorbox}
  - \usepackage{mdframed}
  - \usepackage{geometry}
  - \usepackage{tikz}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Declare Variables

```{r}
num_var <- 10

int_var <- 7L

char_var <- "Bioinformatics"

complex_var <- 4 + 3i

print(num_var)
print(int_var)
print(char_var)
print(complex_var)

```

# Data Type
- return the type of each variable

```{r}
typeof(num_var)
typeof(int_var)
typeof(char_var)
typeof(complex_var)
```

# Countdown using while loop
- countdown from 10 to 0 until the condition is false

```{r}
count <- 10
while (count >= 0) {
  print(count)
  count <- count - 1
}
```

# Function to check even or odd

```{r}
check_even_odd <- function(num) {
  if (num %% 2 == 0) {
    print("Even")
  } else {
    print("Odd")
  }
}
```

# Create a vector
- group a collection of elements together

```{r}
vec <- c(1,2,3,4,5,6,7,8,9,10)

for (element in vec) {
  print(element)
}
```

# Create a 4D array with random numbers

```{r}
array_4d <- array(runif(16, min=0, max=10), dim = c(2,2,2,2))
print(array_4d)
```

# Iris
- we use the flowers data set to perform some operations

```{r}
data(iris)

num_rows <- nrow(iris)
num_cols <- ncol(iris)

column_names <- colnames(iris)

filtered_rows <- subset(iris, Petal.Length > 1.5 & Species == "setosa")

print(paste("Number of rows:", num_rows))
print(paste("Number of columns:", num_cols))
print("Column names:")
print(column_names)
print(paste("Rows where Petal.Length > 1.5 & Species == Setosa:", nrow(filtered_rows)))

```

# Dependency
- we need some libraries to perform some operations optimally

```{r eval=FALSE}
install.packages('tidyverse')
library(tidyverse)
library(dplyr)
```

# Read data-set
- we read the data set from a csv file

```{r}
dataset <- read.csv("BrainCancerMin.csv")

print(paste("-Number of rows =", nrow(dataset)))
print(paste("-Number of columns =", ncol(dataset)))
print("-Column names are")
print(colnames(dataset))
```

# Data pre-processing

##   Determining the Working Set
- here we select the subset of the working data we want to work with
- we select the samples, type, and the first 3 and last 4 genes
- we also count the number of each type of cancer and plot the distribution
- we use ggplot2 to plot the distribution

```{r}
library(dplyr)
subset_dataset <- dataset %>% select(samples, type, 3:5, 147:150)

type_count <- table(subset_dataset$type)
the_most_occurring_type_of_cancer <- names(which.max(type_count))
print(paste("The most occurring type of cancer is:", the_most_occurring_type_of_cancer))

library(ggplot2)
cancer_dataframe <- as.data.frame(type_count)
colnames(cancer_dataframe) <- c("Type", "Count")

ggplot(cancer_dataframe, aes(x = reorder(Type, -Count), y = Count, fill = Type)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Set2") +  # Use different colors for each type
  labs(title = "Cancer Type Distribution",
       x = "Cancer Type",
       y = "Frequency") +
  theme_minimal() +
  theme(legend.position = "none")  # Hide legend if unnecessary
```

##   Data Cleaning and Filtering
- we remove the rows with NA values
- we filter the data set to keep only the rows with  gene X1007_s_at > 12
- we print the number of rows before and after filtering
- we also print the number of NA values in the data set

```{r}
print(paste("-The number of NA in dataset is", sum(is.na(dataset))))

filtered_dataset <- dataset %>% filter(X1007_s_at > 12)
print(paste("-The number of rows before filtering is", nrow(dataset)))
print(paste("-The number of rows after filtering is", nrow(filtered_dataset)))
```

# Data Analysis

##   Genes Analysis
- we calculate the mean and standard deviation of each gene
- we print the results in a new data frame

```{r}
genes <- dataset %>% select(!(1:2))

mean_summary <- summarise(genes, across(where(is.numeric), 
                                          \(x) mean(x, na.rm = TRUE)))

sd_summary <- summarise(genes, across(where(is.numeric),
                                          \(x) sd(x, na.rm = TRUE)))


gene_summary <- bind_rows(mean_summary, sd_summary) %>%
  mutate(Summary = c("mean", "sd")) %>%
  select(Summary, everything())
  
print(gene_summary)

```

##   Genes Analysis By Type
- we calculate the mean and standard deviation of each gene by type
- we print the results in a new data frame
```{r}
library(dplyr)
library(tidyr)
grouped_summary <- dataset %>%
  group_by(type) %>%
  summarise(across(starts_with("X"),
                   list(mean = ~mean(.x, na.rm = TRUE),
                        sd = ~sd(.x, na.rm = TRUE)))) %>%
  pivot_longer(-type, names_to = c("Gene", "Measure"),
               names_pattern = "(.*)_(mean|sd)") %>%
  pivot_wider(names_from = Gene, values_from = value) %>%
  mutate(Measure = paste(Measure, type, sep = "_")) %>%
  select(-type) 

colnames(grouped_summary)[1] <- "measure"

print(grouped_summary)

```

##   Save summaries to csv files
- here we create a function to save the summaries to csv files

```{r}
save_to_csv <- function(ds, path) {
  if(!endsWith(path, ".csv")){
    path <- paste0(path, ".csv")
  }
  
  write.csv(ds, path, row.names = TRUE)
}

save_to_csv(gene_summary, "gene_summary.csv")
save_to_csv(grouped_summary, "grouped_summary.csv")
```